---
title: "Diagnostics and Model Comparison"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Diagnostics and Model Comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.5,
  message  = FALSE,
  warning  = FALSE
)
library(mrmopt)
library(dplyr)
library(ggplot2)
```

Fitting a response curve is the easy part. Knowing whether the fit is
trustworthy requires checking convergence diagnostics, evaluating predictive
performance, and comparing candidate curve types. This vignette covers the
full diagnostic and comparison workflow.

---

## Fit two candidate models

We'll work with the Paid Social channel from `mrmopt_data`, comparing a
Gompertz and a logistic curve:

```{r fit-models, message = FALSE}
social_data <- mrmopt_data |> filter(channel == "Paid Social")

fit_gomp <- fit_response(
  data           = social_data,
  spend          = "spend",
  kpi            = "conversions",
  date           = "week",
  type           = "gompertz",
  midpoint_range = c(0.1, 0.5),
  ceiling_max    = 3,
  refresh        = 0,
  iter = 1000,
  chains = 2
)

fit_logistic <- fit_response(
  data           = social_data,
  spend          = "spend",
  kpi            = "conversions",
  date           = "week",
  type           = "logistic",
  midpoint_range = c(0.1, 0.5),
  ceiling_max    = 3,
  refresh        = 0,
  iter = 1000,
  chains = 2
)
```

---

## Step 1: MCMC convergence

Before interpreting any results, verify that the sampler converged. A model that
hasn't converged can produce arbitrarily wrong parameter estimates.

### Trace plots

`mrm_plot_diagnostics()` produces trace plots for the four curve parameters
plus a posterior predictive check:

```{r diag-gomp, fig.height = 6}
mrm_plot_diagnostics(fit_gomp)
```


::: {.callout}

Note: An `mrmfit` object is a `brms` object so further diagnostics with functions from `bayesplot` will work as expected

:::


```{r}
class(fit_logistic)
```

**What to look for in trace plots:**

- **Good**: Chains overlap completely, forming a "fuzzy caterpillar." No trends,
  no drift, no one chain stuck in a different region.
- **Bad**: Chains that separate (bimodal posterior), chains that trend upward or
  downward (haven't converged), or chains that get stuck in one region for many
  iterations (poor mixing).

### Numerical diagnostics: Rhat and ESS

The `summary()` method reports Rhat and Bulk ESS for each parameter:

```{r summary-gomp}
summary(fit_gomp)
```

**Targets:**

| Diagnostic | Target | Concern |
|------------|--------|---------|
| Rhat | ≤ 1.01 | Values above 1.01 indicate chains haven't converged to the same distribution |
| Bulk ESS | > 400 | Below 400, posterior summaries (means, quantiles) may be unreliable |
| Tail ESS | > 400 | Low tail ESS means credible interval endpoints are unstable |

If Rhat > 1.01 for any parameter, do not trust the results. Increase `iter`
and `warmup`, or adjust priors (see [Priors and Model Tuning](priors_and_tuning.html)).

---

## Step 2: Posterior predictive check

The bottom panel of `mrm_plot_diagnostics()` shows a **posterior predictive
check** (PPC). This compares the observed data distribution to data simulated
from the fitted model.

**What to look for:**

- **Good**: The blue posterior draws envelope the black observed-data line. The
  overall shape and spread match.
- **Bad**: A systematic shift (model predicts higher/lower than observed), a
  shape mismatch (model is unimodal but data is bimodal), or the observed data
  falls outside the posterior draws entirely.

A PPC failure usually means the curve form is wrong for this data, not that the
sampler failed. Try a different curve type.

---

## Step 3: Bayes R²

Bayes R² provides a posterior distribution of explained variance. It is reported
in both `print()` and `summary()`:

```{r r2-compare}
cat("Gompertz R²:\n")
print(fit_gomp)

cat("\nLogistic R²:\n")
print(fit_logistic)
```

**Interpretation:**

- **R² > 0.6**: Well-identified curve. The response model explains a
  substantial share of the variation. Current mix of partners, tactics, audiences within a channel are sufficiently homogeneous to be modeled by one response curve.
- **R² 0.4–0.6**: Moderate fit. The curve captures the general trend but there
  is significant unexplained noise. Variation in the data may suggest the need for more nuanced modeling or classification of channels.
- **R² < 0.4**: Weak fit. The curve may not be well-identified from this data.
  Consider whether the data has enough spend variation, or whether the
  relationship is genuinely noisy. The channel is likely composed of subchannels with distinct response behavior, or is reflective of large strategic shifts in media buy and execution.

Bayes R² is useful for relative comparison between curve types on the same
data, but the absolute value depends heavily on the noise level in the channel. A lower R² does not render the model useless, but does warrant a further investigation into the need for more nuanced modeling or classification of channels.

---

## Step 4: Visual comparison

`mrms_plot_compare()` overlays multiple fitted curves on a single axis,
making shape differences immediately visible:

```{r compare-overlay}
mrms_plot_compare(
  list(gompertz = fit_gomp, logistic = fit_logistic),
  interval = "confidence"
)
```

The `interval` argument controls the uncertainty band:

- `"prediction"` (default): Includes observation noise — wider, represents
  where future data points might fall.
- `"confidence"`: Shows uncertainty in the mean curve only — narrower, better
  for comparing curve shapes.
- `"none"`: Just the fitted curves.

### Faceted layout

For more than two models, a faceted layout can be easier to read:

```{r compare-facet, eval = FALSE}
mrms_plot_compare(
  list(gompertz = fit_gomp, logistic = fit_logistic),
  layout = "facet",
  interval = "prediction"
)
```

### Return curves

Compare marginal and average return curves to see how the curve type affects
efficiency estimates:

```{r compare-return}
mrms_plot_compare(
  list(gompertz = fit_gomp, logistic = fit_logistic),
  plot_type = "return",
  interval = "none"
)
```

Differences in the marginal return curve between curve types directly affect
optimization results — a model that predicts faster diminishing returns will
recommend reallocating spend away from that channel sooner.

---

## Step 5: Comparing across all six curve types

For a thorough model selection, fit all six types and compare:

```{r fit-all-types, message = FALSE}
curve_types <- c(
  "gompertz", "logistic", "reflected_gompertz",
  "log_logistic", "weibull", "reflected_weibull"
)

fits <- lapply(curve_types, function(ct) {
  fit_response(
    data           = social_data,
    spend          = "spend",
    kpi            = "conversions",
    date           = "week",
    type           = ct,
    midpoint_range = c(0.1, 0.5),
    ceiling_max    = 3,
    refresh        = 0,
    iter = 1000,
    chains = 2
  )
})
names(fits) <- curve_types
```

```{r compare-all}
mrms_plot_compare(fits, interval = "confidence")
```

---

## A decision framework

Use this sequence to choose a curve type:

1. **Eliminate non-converged models.** Any model with Rhat > 1.01 or very low
   ESS is unreliable and should not be compared.
2. **Check the PPC.** Models with systematic PPC failures are misspecified for
   this data, regardless of R².
3. **Compare Bayes R².** Among surviving models, prefer higher R² — but treat
   small differences (< 0.02) as noise.
4. **Compare credible band width.** Tighter bands mean the data more strongly
   constrains the curve. A model with slightly lower R² but much tighter bands
   may be preferable.
5. **Check domain plausibility.** Does the ceiling estimate make sense? Is the
   midpoint in a reasonable range? A model that fits the data slightly better
   but produces implausible parameter estimates is less useful for optimization.

When two models are genuinely indistinguishable, prefer the simpler one
(Gompertz or log-logistic over reflected variants) or run the optimization with
both and check whether the allocation decision changes materially.

---

## Diagnostic workflow for a full portfolio

When fitting multiple channels for optimization, run diagnostics on every
channel — not just the first one. A quick check loop:

```{r portfolio-check, eval = FALSE}
for (ch_name in names(fits_portfolio)) {
  cat("\n===", ch_name, "===\n")
  print(summary(fits_portfolio[[ch_name]]))
}
```

If any channel has convergence issues, the optimization results for the
entire portfolio are suspect — the optimizer will misallocate spend to/from
poorly identified channels.

---

## Where to go next

| Topic | Vignette |
|-------|----------|
| Tuning priors when diagnostics reveal problems | [Priors & Model Tuning](priors_and_tuning.html) |
| Budget optimization across channels | [Optimization](optimization.html) |
| Hierarchical models for sub-channel structure | [Hierarchical Models](hierarchical_models.html) |
