---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.5,
  message  = FALSE,
  warning  = FALSE
)
library(mrmopt)
library(dplyr)
library(ggplot2)
```

This vignette walks through the full `mrmopt` workflow from raw data to an
optimized media allocation, using the built-in `mrmopt_data` dataset. By the
end you will have fit response curves for five channels, compared curve types,
checked diagnostics, and produced a budget optimization with uncertainty.

---

## Where mrmopt sits in the analytics stack

`mrmopt` is a **response curve and optimization layer**. It takes
weekly spend and attributed KPI data by channel — the kind of output produced
by an upstream Media Mix Model or Multi-Touch Attribution model — and adds
three things the upstream model typically doesn't provide:

1. **Nonlinear saturation curves** — explicitly modeling diminishing returns
   rather than assuming a linear relationship between spend and KPI
2. **Bayesian uncertainty** — credible intervals on curves, marginal returns,
   and cost-per-unit that honestly reflect what the data does and doesn't tell
   you
3. **Uncertainty-aware optimization** — budget allocation that accounts for
   parameter uncertainty rather than optimizing against a single point estimate

`mrmopt` does not replace a joint MMM. It operates downstream of one.

---

## The workflow

Every analysis in `mrmopt` follows the same iterative loop:

```{r workflow-diagram, echo = FALSE, eval = FALSE}
# Rendered as a Mermaid diagram in pkgdown
```

```
Raw data
  └─► fit_response()          # Fit one channel, one curve type
        └─► mrm_plot_diagnostics()  # Check convergence
              └─► mrms_plot_compare()  # Compare curve types
                    └─► fit_response()  # Best type, full iterations, all channels
                          └─► opt_mix()  # Optimize the portfolio
                                └─► opt_summary() / opt_plot_*()
```

The two loops — convergence checking and curve type selection — are the parts
new users most often skip. Skipping them produces unreliable curves and
misleading optimization results. This vignette shows you what to look for at
each step.

---

## The data

`mrmopt_data` is a built-in dataset with five media channels, each with 104
weeks (two years) of weekly spend and attributed conversions.

```{r data}
data(mrmopt_data)
mrmopt_data
```

```{r data-summary}
mrmopt_data |>
  group_by(channel) |>
  summarise(
    weeks        = n(),
    total_spend  = sum(spend),
    total_conv   = sum(conversions),
    avg_wk_spend = mean(spend),
    avg_wk_conv  = mean(conversions)
  )
```

**Required format:** One row per channel-week. At minimum you need `date`,
`spend`, and `kpi` columns. An optional `units` column (impressions, GRPs,
clicks) unlocks cost-per-unit metrics. Channels are modeled one at a time —
filter to a single channel before calling `fit_response()`.

---

## Step 1: Fit a response curve

Start with one channel and one curve type. Gompertz is a reasonable default
for channels with concave diminishing returns from the start.

```{r fit-ps, message = FALSE, warning = FALSE}
ps_data <- mrmopt_data |> filter(channel == "Paid Search")

fit_ps <- fit_response(
  data           = ps_data,
  spend          = "spend",
  kpi            = "conversions",
  date           = "week",
  type           = "gompertz",
  midpoint_range = c(0.1, 0.5),
  ceiling_max    = 3,
  refresh        = 0,
  iter = 1000,
  chains = 2
)
```

`fit_response()` fits a Bayesian nonlinear model using Stan via `brms`.
`midpoint_range` and `ceiling_max` are scale-invariant prior controls —
you do not need to specify raw Stan priors unless you want to.

---

## Step 2: Read the output

`print()` gives a five-section summary of the fitted model:

```{r print-fit}
print(fit_ps)
```

Reading this top to bottom:

- **Response Curve Summary** — the curve type and the channel/KPI columns used
- **Current Performance** — where the channel sits right now: weekly spend,
  KPI, cost-per-conversion, average return (AR), and marginal return (MR)
- **Parameters** — the four fitted curve parameters. `d` (ceiling) and `e`
  (midpoint) are the most interpretable: ceiling is the maximum predicted KPI,
  midpoint is the spend level where the curve is steepest
- **Response Curve Summary** — three reference points on the curve anchored to
  MR multiples, helping you see how much room is left to grow and how quickly
  returns are declining
- **Bayes R²** — a posterior distribution of explained variance. Higher is
  better; values below 0.4 suggest the curve may not be well-identified from
  this data

`mrm_params()` gives a focused table of just the four parameters:

```{r params}
mrm_params(fit_ps)
```

---

## Step 3: Check diagnostics

Before interpreting anything, confirm the model converged.

```{r diagnostics, fig.height = 6}
mrm_plot_diagnostics(fit_ps)
```

Two things to check:

- **Trace plots** — four chains should mix freely (overlap completely, no
  trends or drifts). If chains are separating or stuck, the model has
  convergence problems.
- **Posterior predictive check** — the blue posterior draws should envelope
  the black observed data line. A systematic gap means the curve form is
  misspecified.

The numerical indicators to target are **Rhat ≤ 1.01** and **Bulk ESS > 400**
for all parameters. These are printed in `mrm_summary()`:

```{r mrm-summary}
summary(fit_ps)
```

If convergence is poor, try increasing `iter` and `warmup`, or adjust the
priors with a tighter `midpoint_range`.

---

## Step 4: Visualize the response curve

Once convergence is confirmed, look at the curve itself:

```{r plot-fit}
mrm_plot(fit_ps)
```

This dashboard shows three panels:

- **Response curve** — KPI vs. spend with posterior credible band. The current
  operating point is marked.
- **Average & marginal return** — AR and MR curves. Where MR is well below AR,
  the channel is in the diminishing returns zone. Where MR ≈ AR, it is
  underinvested.
- **Cost-per** — cost-per-conversion across the spend range. Flat or declining
  cost-per means the channel absorbs more spend efficiently.

---

## Step 5: Compare curve types

`mrmopt` supports six curve forms. The right choice depends on the data — some
channels have S-shaped response (low returns at very low spend, peak efficiency
in the middle, diminishing returns at high spend) while others show concave
diminishing returns from the first dollar.

Fit a second type and compare them visually:

```{r fit-compare, message = FALSE, warning = FALSE}
fit_ps_ll <- fit_response(
  data           = ps_data,
  spend          = "spend",
  kpi            = "conversions",
  date           = "week",
  type           = "log_logistic",
  midpoint_range = c(0.1, 0.5),
  ceiling_max    = 3,
  refresh        = 0,
  iter = 1000,
  chains = 2
)
```

```{r compare-plot}
mrms_plot_compare(
  list(gompertz = fit_ps, log_logistic = fit_ps_ll),
  interval = "predictio"
)
```

Prefer the curve that:

1. Has tighter credible bands (better identified from the data)
2. Has higher Bayes R²
3. Shows no systematic gap between the predictive band and the data in the
   diagnostic PPC

For a full comparison across all six types, see the
[Diagnostics & Comparison](diagnostics_and_comparison.html) vignette.

---

## Step 6: Fit all channels

Once you have a curve type for each channel, fit the full portfolio. Here we
use the best-fit type for each of the five channels in `mrmopt_data`:

```{r fit-all, message = FALSE, warning = FALSE}
channel_types <- list(
  "Paid Search"  = "log_logistic",
  "Paid Social"  = "gompertz",
  "Display"      = "gompertz",
  "Online Video" = "gompertz",
  "TV"           = "gompertz"
)

fits <- lapply(names(channel_types), function(ch) {
  fit_response(
    data  = mrmopt_data |> filter(channel == ch),
    spend = "spend",
    kpi   = "conversions",
    date  = "week",
    type  = channel_types[[ch]],
    midpoint_range = c(0.1, 0.5),
    ceiling_max    = 3,
    refresh        = 0,
    iter = 1000,
    chains = 2
  )
})
names(fits) <- names(channel_types)
```

Compare all five on the same axes to get an immediate sense of relative
efficiency:

```{r compare-all}
mrms_plot_compare(fits, interval = "confidence")
```

---

## Step 7: Optimize the portfolio

`opt_mix()` finds the spend allocation that maximizes total KPI subject to a
total budget constraint. The default `method = "point"` uses the posterior
median parameters and runs in under a second:

```{r opt, message = FALSE, warning = FALSE}
opt <- opt_mix(fits, budget = 204641)
```

`opt_summary()` prints a formatted allocation table:

```{r opt-summary}
opt_summary(opt)
```

Reading the output:

- **Weekly Spend** — the optimal spend per channel at the given budget
- **Weekly KPI** — predicted conversions at the optimal spend
- **CP** — cost-per-conversion at the optimal point
- **Share** — share of total budget
- **Totals** — current vs. optimal KPI and cost-per, with the overall change

`opt_table()` returns the same information as a tidy tibble for further
analysis or reporting:

```{r opt-table}
opt_table(opt) |>
  select(channel, current_spend, optimal_spend, spend_delta_pct,
         current_kpi, optimal_kpi, kpi_delta_pct, cp_delta)
```

---

## Step 8: Visualize the allocation

Two plot functions cover the most common views:

```{r plot-alloc}
opt_plot_allocation(opt)
```

```{r plot-comparison}
opt_plot_comparison(opt)
```

To see exactly where the optimal spend falls on each channel's response curve:

```{r plot-curves, fig.height = 6}
opt_plot_curves(opt)
```

And on the marginal return curves — the view that shows *why* the optimizer
moved spend the way it did:

```{r plot-returns, fig.height = 6}
opt_plot_returns(opt)
```

---

## Step 9: Add uncertainty with posterior optimization

The point estimate result is fast but treats the posterior median parameters as
if they were known exactly. `method = "posterior"` runs the optimizer over a
sample of posterior draws, producing a distribution of optimal allocations:

```{r opt-post, message = FALSE, warning = FALSE}
opt_post <- opt_mix(fits, budget = 204641, method = "posterior", n_draws = 200)
```

```{r opt-post-summary}
opt_summary(opt_post)
```

The `[95% CI]` column shows the range of optimal spend across draws.
`opt_plot_allocation()` adds error bars:

```{r plot-post-alloc}
opt_plot_allocation(opt_post)
```

And `opt_plot_posterior()` shows the full spend distribution per channel:

```{r plot-post-posterior}
opt_plot_posterior(opt_post)
```

Wide distributions signal channels where the data does not strongly constrain
the curve — the optimizer is uncertain about where to put spend because the
response is uncertain. Narrow distributions signal well-identified channels
where the allocation decision is robust.

---

## Scenario analysis

`opt_mix()` supports several common scenario needs:

**Period budgets** — optimize a quarterly or annual budget broken into weekly
allocations:

```{r opt-annual, message = FALSE, warning = FALSE}
opt_annual <- opt_mix(fits, budget = 26000000, n_weeks = 52)
```

**Custom constraints** — set channel-level floors, ceilings, and share bounds:

```{r opt-constrained, message = FALSE, warning = FALSE}
constraints <- data.frame(
  channel   = names(fits),
  min_spend = c(10000, 5000, 2000, 5000, 50000),
  max_spend = c(200000, 150000, 50000, 100000, 300000)
)

opt_constrained <- opt_mix(fits, budget = 204641, constraints = constraints)
```

See the [Optimization](optimization.html) vignette for the full constraint
specification including share bounds and fixed channels.

---

## Where to go next

| Topic | Vignette |
|-------|----------|
| How priors work and when to adjust them | [Priors & Model Tuning](priors_and_tuning.html) |
| Convergence checks and model selection | [Diagnostics & Model Comparison](diagnostics_and_comparison.html) |
| All optimization objectives, constraints, and methods | [Optimization](optimization.html) |
| Sub-channel hierarchical models | [Hierarchical Models](hierarchical_models.html) |
| Mathematical derivations of all six curve forms | [Response Curve Theory](response_curve_theory.html) |
