---
title: "Hierarchical Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hierarchical Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.5,
  message  = FALSE,
  warning  = FALSE
)
library(mrmopt)
library(dplyr)
library(ggplot2)
```

Media channels are rarely homogeneous. TV spend spans broadcast, cable, and
streaming — each with different audience reach and saturation dynamics. Social
covers multiple partners, tactics, and creatives. Standard response curve
modeling treats the channel as a single unit, which either overfits sparse
sub-channel data or discards useful granularity.

`fit_response_hier()` fits a single hierarchical model where curve parameters
are **partially pooled** across sub-channel groupings. Sparse units borrow
strength from better-identified peers, while well-observed units retain their
individual character. This vignette walks through the full hierarchical
workflow using the built-in `mrmopt_hier_data` dataset.

---

## The data

`mrmopt_hier_data` contains weekly spend and conversions for eight TV partners
nested within three subtypes:

```{r data}
data(mrmopt_hier_data)
mrmopt_hier_data
```

```{r data-summary}
mrmopt_hier_data |>
  group_by(subtype, partner) |>
  summarise(
    weeks       = n(),
    avg_spend   = mean(spend),
    avg_conv    = mean(conversions),
    .groups     = "drop"
  )
```

The hierarchy has two levels: `subtype` (Broadcast, Cable, Streaming) and
`partner` (the individual units within each subtype). Broadcast partners have
higher spend and ceiling than Streaming partners, but all share a common curve
family.

```{r data-plot, fig.height = 5}
ggplot(mrmopt_hier_data, aes(x = spend, y = conversions)) +
  geom_point(alpha = 0.4, size = 1) +
  facet_wrap(~partner, scales = "free") +
  scale_x_continuous(labels = scales::dollar_format(scale = 1e-3, suffix = "k")) +
  labs(x = "Weekly Spend", y = "Conversions") +
  theme_minimal()
```

---

## When to use hierarchical models

Use `fit_response_hier()` when:

- A channel has **natural sub-groups** (partners, tactics, creatives, regions)
  that share a common response shape but differ in scale
- Some sub-groups have **sparse data** — too few weeks to identify a curve on
  their own, but enough to contribute information when pooled
- You want **per-unit response curves** for optimization or reporting, but don't
  want to fit each unit independently and risk overfitting

Do not use hierarchical models when:

- Sub-groups have fundamentally different response dynamics (different curve
  *types*, not just different parameters). Fit them separately.
- You have only one grouping level with 2 groups — the pooling SD will be
  poorly identified.
- All sub-groups have ample data and you don't need borrowing. Independent fits
  via `fit_response()` are simpler and faster.

---

## Fit the hierarchical model

Specify the grouping columns in order from broadest to finest:

```{r fit-hier, message = FALSE}
fit_tv <- fit_response_hier(
  data           = mrmopt_hier_data,
  spend          = "spend",
  kpi            = "conversions",
  date           = "week",
  group          = c("subtype", "partner"),
  type           = "gompertz",
  pool           = c("b", "e", "d"),
  midpoint_range = c(0.1, 0.6),
  ceiling_max    = 3,
  refresh        = 0,
  iter = 1000,
  chains = 2
)
```

The `group` argument defines the nesting structure. With `group = c("subtype",
"partner")`, the model fits:

- **Fixed effects**: The channel-level (population) mean curve
- **Random effects at level 1** (`subtype`): Deviations for Broadcast, Cable,
  Streaming
- **Random effects at level 2** (`subtype:partner`): Partner-level deviations
  nested within subtypes

The `pool` argument controls which parameters receive random effects. The
default `c("b", "e", "d")` pools steepness, midpoint, and ceiling. The floor
`c` is shared across all units — it represents baseline conversions at zero
spend, which is usually a channel-level property.

---

## Diagnostics

Check convergence the same way as a single-channel model:

```{r diag-hier, fig.height = 6}
mrm_plot_hier_diagnostics(fit_tv)
```

Look for the same targets: clean trace plots, Rhat ≤ 1.01, ESS > 400. The
hierarchical model has more parameters (population + group SDs + per-unit
deviations), so convergence can be slower. If you see issues, try:

- Increasing `iter` and `warmup`
- Reducing the number of pooled parameters (e.g., `pool = c("b", "e")`)
- Increasing `adapt_delta` via `control = list(adapt_delta = 0.99)`

---

## Summaries

`print()` provides a concise overview:

```{r print-hier}
print(fit_tv)
```

`mrm_summary_hier()` produces a detailed summary with one row per unit at every
hierarchy level, plus a channel-level aggregate:

```{r summary-hier}
mrm_summary_hier(fit_tv)
```

Each row shows the fitted curve parameters, current performance, and the
efficient operating range (range_min to range_max) for that unit.

---

## Visualization

### Response curves by level

The dashboard view shows response curves at each hierarchy level:

```{r plot-hier-dashboard, fig.height = 8}
mrm_plot_hier(fit_tv)
```

For response curves at a specific level:

```{r plot-hier-response, fig.height = 5}
mrm_plot_hier_response(fit_tv)
```

### Shrinkage

The shrinkage plot shows how each unit's parameter estimate is pulled toward the
group mean by partial pooling. Units with more data are pulled less; units with
sparse data are pulled more:

```{r plot-shrinkage}
mrm_plot_hier_shrinkage(fit_tv, param = "d")
```

This is the core benefit of hierarchical modeling — sparse partners get
regularized estimates rather than noisy independent fits.

---

## Per-unit optimization

To optimize across sub-channel units, convert the hierarchical fit to a list of
single-curve model views using `as_mrmfit_list()`:

```{r as-list}
unit_fits <- as_mrmfit_list(fit_tv)
names(unit_fits)
```

Each element is an `mrmfit_hier_unit` object that behaves like a regular
`mrmfit` for optimization purposes. Pass the list to `opt_mix()`:

```{r opt-units}
opt_units <- opt_mix(unit_fits, budget = 130000)
opt_summary(opt_units)
```

The optimizer now allocates spend across all eight partners simultaneously,
respecting each unit's response curve:

```{r opt-plot, fig.height = 6}
opt_plot_curves(opt_units)
```

### Optimizing at a higher level

By default, `as_mrmfit_list()` extracts at the innermost (finest) level. To
optimize at the subtype level instead:

```{r opt-subtype}
subtype_fits <- as_mrmfit_list(fit_tv, level = "subtype")
names(subtype_fits)
```

```{r opt-subtype-result}
opt_subtypes <- opt_mix(subtype_fits, budget = 130000)
opt_summary(opt_subtypes)
```

This is useful when you control budget at the subtype level but want the
subtype-level curves to reflect partial pooling from the partner data below.

---

## The `pool` parameter

The choice of which parameters to pool affects how much information is shared:

| Setting | Effect |
|---------|--------|
| `pool = c("b", "e", "d")` | Default. Shape and scale vary by unit. Most flexible. |
| `pool = c("b", "e")` | Shape varies; ceiling `d` is fixed at channel level. Use when units have similar market size. |
| `pool = c("d")` | Only scale varies; shape is shared. Use when units have similar saturation dynamics but different sizes. |
| `pool = c("b", "c", "d", "e")` | All four parameters pooled. Maximum borrowing. |

Adding more pooled parameters increases model complexity and can slow
convergence. Start with the default and remove parameters from `pool` if
convergence is poor or if domain knowledge says certain parameters should be
shared.

---

## Where to go next

| Topic | Vignette |
|-------|----------|
| Budget optimization objectives and constraints | [Optimization](optimization.html) |
| Prior tuning when hierarchical fits struggle | [Priors & Model Tuning](priors_and_tuning.html) |
| Mathematical foundations of the six curve forms | [Response Curve Theory](response_curve_theory.html) |
