---
title: "Optimization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Optimization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.5,
  message  = FALSE,
  warning  = FALSE
)
library(mrmopt)
library(dplyr)
library(ggplot2)
```

`opt_mix()` takes a portfolio of fitted response curves and finds the spend
allocation that achieves a specified objective — maximizing KPI, hitting a ROI
target, or reaching a marginal efficiency threshold. This vignette covers all
five objectives, the constraint system, point vs. posterior methods, and how
to interpret and visualize results.

---

## Setup: fit the portfolio

We'll use all five channels from `mrmopt_data`. In practice, you should have
already run diagnostics on each fit (see
[Diagnostics & Model Comparison](diagnostics_and_comparison.html)).

```{r fit-all, message = FALSE}
channel_types <- list(
  "Paid Search"  = "log_logistic",
  "Paid Social"  = "gompertz",
  "Display"      = "gompertz",
  "Online Video" = "gompertz",
  "TV"           = "gompertz"
)

fits <- lapply(names(channel_types), function(ch) {
  fit_response(
    data  = mrmopt_data |> filter(channel == ch),
    spend = "spend",
    kpi   = "conversions",
    date  = "week",
    type  = channel_types[[ch]],
    midpoint_range = c(0.1, 0.5),
    ceiling_max    = 3,
    refresh        = 0,
    iter = 1000,
    chains = 2
  )
})
names(fits) <- names(channel_types)
```

The current total weekly spend across all channels:

```{r current-budget}
current_budget <- sum(sapply(fits, function(f) {
  mean(f$data[[f$spend_col]])
}))
cat("Current weekly budget: $", scales::comma(round(current_budget)), "\n")
```

---

## Objective 1: Maximize KPI (fixed budget)

The default objective. Given a total budget, find the allocation that maximizes
total predicted KPI:

```{r opt-max-kpi}
opt <- opt_mix(fits, budget = 200000)
```

`opt_summary()` prints a formatted allocation table:

```{r opt-summary}
opt_summary(opt)
```

`opt_table()` returns the same information as a tidy tibble, useful for
downstream analysis or reporting:

```{r opt-table}
opt_table(opt) |>
  select(channel, current_spend, optimal_spend, spend_delta_pct,
         current_kpi, optimal_kpi, kpi_delta_pct)
```

### Visualizing the result

```{r plot-allocation}
opt_plot_allocation(opt)
```

The dumbbell chart shows the magnitude and direction of each reallocation:

```{r plot-comparison}
opt_plot_comparison(opt)
```

To see where each channel's optimal spend falls on its response curve:

```{r plot-curves, fig.height = 6}
opt_plot_curves(opt)
```

And on the marginal/average return curves — the view that explains *why* the
optimizer moved spend the way it did:

```{r plot-returns, fig.height = 6}
opt_plot_returns(opt)
```

At the optimum, marginal returns are equalized across channels (within
constraint bounds). Channels where the optimal point has higher MR than others
are constrained — the optimizer would like to allocate more but is hitting a
bound.

---

## Objective 2: Target ROI (flexible budget)

Instead of fixing the budget and maximizing KPI, target a minimum portfolio
ROI and let the optimizer find the budget:

```{r opt-roi}
opt_roi <- opt_mix(fits, objective = "target_roi", target_roi = 2.0)
opt_summary(opt_roi)
```

The budget is now an *output*, not an input:

```{r roi-budget}
cat("Weekly budget at target ROI:", scales::dollar(opt_roi$budget_info$weekly_budget), "\n")
cat("Achieved ROI:", round(opt_roi$achieved_roi, 2), "\n")
```

ROI here is defined as total incremental KPI divided by total spend:

$$\text{ROI} = \frac{\sum_i [f_i(x_i) - f_i(0)]}{\sum_i x_i}$$

where $$f_i(0)$$ is the floor (baseline KPI at zero spend) for each channel.

---

## Objective 3: Target marginal ROI (per-channel)

Target mROI sets each channel's spend to where its marginal return equals a
threshold. This is a per-channel operation — no cross-channel optimization is
needed:

```{r opt-mroi}
opt_mroi <- opt_mix(fits, objective = "target_mroi", target_mroi = 0.01)
opt_summary(opt_mroi)
```

The resulting budget is the sum of per-channel solutions:

```{r mroi-budget}
cat("Weekly budget at target mROI:", scales::dollar(opt_mroi$budget_info$weekly_budget), "\n")
```

Target mROI is useful for setting a "minimum efficiency" threshold: stop
investing in a channel when the next dollar generates less than `target_mroi`
units of KPI.

---

## Objectives 4 & 5: Cost-per-KPI framing

For non-revenue KPIs (leads, visits, sign-ups), thinking in ROI (KPI/spend) is
less natural than cost-per (spend/KPI). The `target_cpk` and `target_mcpk`
objectives are convenience wrappers:

```{r opt-cpk}
# "Stop spending when each conversion costs more than $100 on average"
opt_cpk <- opt_mix(fits, objective = "target_cpk", target_cpk = 100)
opt_summary(opt_cpk)
```

```{r opt-mcpk}
# "Stop spending when the next conversion costs more than $150"
opt_mcpk <- opt_mix(fits, objective = "target_mcpk", target_mcpk = 150)
opt_summary(opt_mcpk)
```

Internally, `target_cpk` converts to `target_roi = 1 / target_cpk` and
`target_mcpk` converts to `target_mroi = 1 / target_mcpk`.

---

## Point vs. posterior optimization

All objectives support two methods:

### Point estimate (default)

Uses the posterior median parameters for each channel. Fast (< 1 second) and
produces a single optimal allocation. Good for quick scenario analysis:

```{r point-method}
opt_point <- opt_mix(fits, budget = 200000, method = "point")
```

### Posterior

Runs the optimizer over `n_draws` posterior draws, producing a *distribution*
of optimal allocations that reflects parameter uncertainty:

```{r posterior-method}
opt_post <- opt_mix(fits, budget = 200000, method = "posterior", n_draws = 100)
opt_summary(opt_post)
```

The `[95% CI]` column shows the range of optimal spend across draws. Posterior
allocation plots include error bars:

```{r plot-post-alloc}
opt_plot_allocation(opt_post)
```

`opt_plot_posterior()` shows the full spend distribution per channel:

```{r plot-post-posterior}
opt_plot_posterior(opt_post)
```

**When to use posterior**: When the allocation decision matters and you want to
understand how sensitive it is to parameter uncertainty. Wide distributions
signal channels where the data doesn't strongly constrain the optimal spend.

**When point is sufficient**: Scenario analysis, quick comparisons, or when all
channels have tight credible bands.

---

## Period budgets

By default, `opt_mix()` optimizes a single-week budget. To optimize an annual
or quarterly budget broken into weekly allocations, use `n_weeks`:

```{r period-budget}
opt_annual <- opt_mix(fits, budget = 10000000, n_weeks = 52)
opt_summary(opt_annual)
```

The optimizer finds the optimal *weekly* allocation, then scales to the period.
The `$solution` tibble contains both weekly and period columns.

---

## Constraints

### Auto-generated (default)

When no `constraints` argument is provided, `opt_mix()` generates bounds
automatically from each channel's fitted return rate ranges, scaled by
`bounds_multiplier` (default 3). This prevents the optimizer from extrapolating
far beyond observed spend levels.

### User-supplied constraints

Pass a data frame with channel-level bounds:

```{r constraints-basic}
my_constraints <- data.frame(
  channel   = names(fits),
  min_spend = c(10000, 5000, 2000, 5000, 50000),
  max_spend = c(100000, 80000, 30000, 60000, 150000)
)

opt_constrained <- opt_mix(fits, budget = 200000, constraints = my_constraints)
opt_summary(opt_constrained)
```

### Share-based constraints

For `max_kpi` (fixed budget) optimization, you can also set minimum and maximum
share-of-budget bounds. When both absolute and share constraints are present,
the tighter bound wins:

```{r constraints-share}
share_constraints <- data.frame(
  channel   = names(fits),
  min_spend = c(10000, 5000, 2000, 5000, 50000),
  max_spend = c(100000, 80000, 30000, 60000, 150000),
  min_share = c(0.05, 0.03, 0.01, 0.03, 0.20),
  max_share = c(0.50, 0.40, 0.15, 0.30, 0.60)
)

opt_shares <- opt_mix(fits, budget = 200000, constraints = share_constraints)
opt_summary(opt_shares)
```

Share-based constraints are only supported for `max_kpi` (fixed budget). They
are warned about and stripped for flexible-budget objectives, since there is
no fixed budget to compute shares against.

### Fixed channels

Lock a channel at its current spend with the `fixed` column:

```{r constraints-fixed}
fixed_constraints <- data.frame(
  channel   = names(fits),
  min_spend = c(10000, 5000, 2000, 5000, 50000),
  max_spend = c(100000, 80000, 30000, 60000, 150000),
  fixed     = c(FALSE, FALSE, FALSE, FALSE, TRUE)
)

opt_fixed <- opt_mix(fits, budget = 200000, constraints = fixed_constraints)
opt_summary(opt_fixed)
```

When `fixed = TRUE`, the channel's spend is locked at `min_spend` and the
remaining budget is optimized across the other channels.

---

## Interpreting the solution

### The solution tibble

The core output is `opt$solution`, a tibble with one row per channel:

```{r solution-cols}
names(opt$solution)
```

Key column groups:

- **Current state** (`current_weekly_*`): Where each channel sits now (from the
  fitted model's data)
- **Optimal state** (`weekly_*`): The optimizer's recommended allocation. For
  posterior results, includes `_lower` and `_upper` CI columns.
- **Period totals** (`period_*`): Weekly values × `n_weeks`
- **Shares** (`spend_share`, `kpi_share`): Fraction of total budget and total
  KPI

### Response rate and cost-per

The solution includes the response rate (`rr`: KPI per dollar) and cost-per
(`cost_per`: dollars per KPI unit) at both the current and optimal points. These
are directly readable from the response curve — they don't require a separate
calculation.

---

## Where to go next

| Topic | Vignette |
|-------|----------|
| Diagnostics to verify fits before optimizing | [Diagnostics & Model Comparison](diagnostics_and_comparison.html) |
| Tuning priors when fits look wrong | [Priors & Model Tuning](priors_and_tuning.html) |
| Hierarchical models for sub-channel optimization | [Hierarchical Models](hierarchical_models.html) |
