---
title: "Priors and Model Tuning"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Priors and Model Tuning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.5,
  message  = FALSE,
  warning  = FALSE
)
library(mrmopt)
library(dplyr)
library(ggplot2)
```

Bayesian models require prior distributions on every parameter. Poor priors can
produce curves that converge numerically but are scientifically implausible —
ceiling estimates ten times the observed maximum, or midpoints far outside the
data range. This vignette explains how `mrmopt`'s prior system works and how to
tune it when the default fit is unsatisfactory.

---

## The three-tier prior system

`mrmopt` offers three levels of prior control. Most users should work at the
simplified level.

### 1. Automatic (default)

When `auto = TRUE` (the default in `fit_response()`), priors are generated
automatically based on the observed data range. This is a reasonable starting
point but uses wide defaults that may be too permissive for some channels.

### 2. Simplified via `mrmopt_prior()`

The recommended approach for most workflows. `mrmopt_prior()` lets you
constrain the four curve parameters using intuitive, scale-invariant inputs:

```{r prior-obj}
my_prior <- mrmopt_prior(
  midpoint_range = c(0.2, 0.6),
  ceiling_max    = 2.5,
  floor_min      = 0,
  anchor_strength = 0.05
)
my_prior
```

These translate to Stan priors internally — you do not need to know the scaled
parameter space.

### 3. Manual via `brms::prior()`

For advanced users who want full control. Pass a raw `brmsprior` object to the
`prior` argument of `fit_response()`. When doing this with `auto = FALSE` and
`scale_data = FALSE`, you take full responsibility for the prior specification.

---

## What each prior control does

### `midpoint_range` — where the inflection falls

A two-element vector giving the lower and upper bound for the inflection point
as a fraction of the x-axis (spend) range. For example, `c(0.2, 0.6)` means
the inflection must fall between the 20th and 60th percentile of observed spend.

This is the most impactful prior control. A tight `midpoint_range` anchors the
curve's shape to a plausible region:

- **Too wide** (e.g., `c(0.01, 0.99)`): The model may place the inflection far
  outside the observed data, producing a nearly linear fit.
- **Too narrow** (e.g., `c(0.3, 0.35)`): Overly informative — the data has
  little room to speak.
- **Good default**: `c(0.1, 0.5)` for channels where you expect diminishing
  returns to be visible within the observed spend range. Widen toward
  `c(0.1, 0.9)` if unsure.

```{r fit-midpoint-tight, message = FALSE}
ps_data <- mrmopt_data |> filter(channel == "Paid Search")

fit_tight <- fit_response(
  data           = ps_data,
  spend          = "spend",
  kpi            = "conversions",
  date           = "week",
  type           = "gompertz",
  midpoint_range = c(0.2, 0.5),
  ceiling_max    = 3,
  refresh        = 0,
  iter = 1000,
  chains = 2
)
```

```{r fit-midpoint-wide, message = FALSE}
fit_wide <- fit_response(
  data           = ps_data,
  spend          = "spend",
  kpi            = "conversions",
  date           = "week",
  type           = "gompertz",
  midpoint_range = c(0.01, 0.99),
  ceiling_max    = 5,
  refresh        = 0,
  iter = 1000,
  chains = 2
)
```

```{r compare-midpoint}
mrms_plot_compare(
  list("tight (0.2–0.5)" = fit_tight, "wide (0.01–0.99)" = fit_wide),
  interval = "confidence"
)
```

The tighter prior produces a more clearly identified curve. The wider prior
allows the credible band to expand substantially, particularly at the extremes
of the spend range.

---

### `ceiling_max` — how high the curve can reach

A multiplier on the observed maximum of the KPI column. `ceiling_max = 3` means
the ceiling parameter `d` can be at most 3× the highest observed weekly
conversions.

- **Too low** (e.g., `1.1`): Forces the ceiling close to the observed max,
  which may be appropriate if you believe the channel is near saturation, but
  can produce a poor fit if the data doesn't clearly plateau.
- **Too high** (e.g., `10`): Allows implausibly large ceilings. The posterior
  may settle on a high ceiling with a far-off midpoint, producing a curve that
  looks linear in-sample.
- **Good default**: `2` to `3` for most channels. Lower if the data clearly
  shows a plateau; higher for channels where you believe substantial headroom
  remains.

---

### `floor_min` — baseline KPI at zero spend

The lower asymptote `c` in original units. Defaults to `0`, meaning the model
assumes zero conversions at zero spend. Set this higher if you have organic
conversions that persist without paid media.

```{r floor-example, eval = FALSE}
# Channel with known organic baseline of ~50 conversions/week
fit_response(
  ...,
  floor_min = 40   # allow floor between 40 and observed data
)
```

---

### `anchor_strength` — how tightly the floor is held

Controls the prior SD on the floor parameter `c` as a fraction of the observed
y range. The default `0.05` (5% of the range) keeps the floor tightly
constrained around `floor_min`, which prevents the curve from developing an
unrealistically high or negative baseline.

- **Smaller values** (e.g., `0.02`): Very tight floor constraint. Use when you
  have strong knowledge of the baseline.
- **Larger values** (e.g., `0.2`): Loose constraint. The floor can move
  substantially from `floor_min`.
- **`NULL`**: No floor constraint beyond the default. Use when you genuinely
  don't know the baseline.

---

## Passing prior controls to `fit_response()`

You can pass `midpoint_range`, `ceiling_max`, `floor_min`, and
`anchor_strength` directly as arguments to `fit_response()` — there is no need
to create an `mrmopt_prior()` object first:

```{r fit-direct, eval = FALSE}
fit_response(
  data           = ps_data,
  spend          = "spend",
  kpi            = "conversions",
  date           = "week",
  type           = "gompertz",
  midpoint_range = c(0.15, 0.5),
  ceiling_max    = 2.5,
  floor_min      = 0,
  anchor_strength = 0.03,
  refresh        = 0
)
```

The standalone `mrmopt_prior()` constructor is useful when you want to inspect
the prior object or reuse the same specification across multiple fits.

---

## Common tuning scenarios

### The curve looks nearly linear

**Symptom**: The response curve is a gentle slope with no visible saturation,
and the ceiling estimate is very high relative to observed data.

**Diagnosis**: The model placed the midpoint far to the right of the observed
data. The curve is in its early accelerating phase across the entire data range.

**Fix**: Tighten `midpoint_range` to force the inflection into the observed
range. Lower `ceiling_max` to prevent runaway ceiling estimates.

```{r fix-linear, eval = FALSE}
fit_response(..., midpoint_range = c(0.1, 0.4), ceiling_max = 2)
```

### Wide credible bands

**Symptom**: The posterior band is so wide it's not informative. The curve
could be nearly flat or steeply rising — the data doesn't distinguish.

**Diagnosis**: Usually indicates insufficient spend variation (see
[Data Requirements](response_curve_theory.html#data-requirements)) or overly
permissive priors.

**Fix**: First check that the data has adequate spend variation. If so, tighten
`midpoint_range` and `ceiling_max`. If not, more data is needed — tighter
priors will reduce uncertainty cosmetically but not substantively.

### The floor is negative or unreasonably high

**Symptom**: The fitted `c` parameter is negative (implying negative KPI at
zero spend) or much higher than expected organic baseline.

**Fix**: Set `floor_min` to a sensible value and reduce `anchor_strength`:

```{r fix-floor, eval = FALSE}
fit_response(..., floor_min = 0, anchor_strength = 0.02)
```

### Divergent transitions or low ESS

**Symptom**: Stan warnings about divergent transitions, or Bulk ESS < 400.

**Diagnosis**: The posterior geometry is difficult for the sampler. This can
happen when priors are contradictory, when the data is very sparse, or when a
parameter is weakly identified.

**Fix**: Try in order:

1. Increase `iter` and `warmup` (e.g., `iter = 4000, warmup = 2000`)
2. Tighten priors to reduce the posterior's effective volume
3. Try a different curve type — some curve forms fit certain data shapes more
   naturally
4. Increase `adapt_delta` via `control = list(adapt_delta = 0.99)`

---

## Inspecting the effective priors

To see the actual Stan priors that `mrmopt` generates, fit a model and examine
the `brms` prior specification:

```{r inspect-priors}
brms::prior_summary(fit_tight)
```

This shows the priors on the scaled parameter space. The mapping between the
simplified controls and these Stan priors depends on the data scaling, which is
why working with the scale-invariant `mrmopt_prior()` interface is recommended.

---

## Where to go next

| Topic | Vignette |
|-------|----------|
| Convergence diagnostics and model comparison | [Diagnostics & Model Comparison](diagnostics_and_comparison.html) |
| Mathematical foundations of the six curve forms | [Response Curve Theory](response_curve_theory.html) |
| Budget optimization across channels | [Optimization](optimization.html) |
